---
title: "ExploreModelMatrix"
author: "Charlotte Soneson"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{exploreModelMatrix}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`ExploreModelMatrix` is an R package for visualizing design matrices generated
by the `model.matrix` R function. Provided with a sample information table and a
design formula, the `exploreModelMatrix()` function launches a shiny app where
the user can explore the modeled values (in terms of the model coefficients) for
each combination of predictor values. In addition, the app allows the user to
interactively change the design formula and the reference levels of factor
variables as well as drop unwanted columns from the design matrix, in order to
explore the effect on the modeled values.

In addition to the interactive visualization, `ExploreModelMatrix` also provides
a function, `visualizeDesign()`, for generating static visualizations.

In this vignette, we show examples of applying the `ExploreModelMatrix`
functions to various experimental design setups. Most examples are taken from
questions raised at the Bioconductor support site.

```{r setup}
library(ExploreModelMatrix)
```

## Example 1 - two categorical predictors

```{r, fig.width = 7}
(sampleData <- data.frame(genotype = rep(c("A", "B"), each = 4),
                          treatment = rep(c("ctrl", "trt"), 4)))
vd <- visualizeDesign(sampleData = sampleData, 
                      designFormula = ~ genotype + treatment, textSize = 4)
cowplot::plot_grid(plotlist = vd$plotlist)
app <- exploreModelMatrix(sampleData = sampleData,
                          designFormula = ~ genotype + treatment)
if (interactive()) shiny::runApp(app)
```

## Example 2 - one continuous predictor

```{r, fig.height = 5}
set.seed(1)
(sampleData <- data.frame(
  age = round(runif(10, min = 12, max = 60))
))
vd <- visualizeDesign(sampleData = sampleData,
                      designFormula = ~ age)
cowplot::plot_grid(plotlist = vd$plotlist, ncol = 1)
```

## Example 3

From https://support.bioconductor.org/p/122506/,
https://support.bioconductor.org/p/122613/

```{r, fig.height = 5, fig.width = 5}
(sampleData <- data.frame(
  group = factor(rep(1:4, each = 6)),
  subject = rep(1:8, each = 3),
  treatment = factor(rep(c("a", "b", "a", "b"), each = 6)),
  time = factor(c(rep(1:3, 4), rep(c(1, 4, 5), 4)))))
vd <- visualizeDesign(sampleData = sampleData,
                      designFormula = ~ 0 + treatment + time:treatment, 
                      flipCoord = TRUE, textSize = 4)
cowplot::plot_grid(plotlist = vd$plotlist)
vd <- visualizeDesign(sampleData = sampleData,
                      designFormula = ~ treatment * time, 
                      flipCoord = TRUE, textSize = 4)
cowplot::plot_grid(plotlist = vd$plotlist)
app <- exploreModelMatrix(sampleData = sampleData,
                          designFormula = ~ 0 + treatment + time:treatment)
if (interactive()) shiny::runApp(app)
```

## Example 4
From https://support.bioconductor.org/p/112947/

```{r, fig.width = 7}
(sampleData = data.frame(
  group = factor(rep(c("NR", "NT", "P", "F"), c(11, 26, 14, 4))),
  batch = factor(c(3, 3, 3, 4, 4, 6, 3, 3, 3, 6, 6, 6, 6, 6, 4, 6, 3, 3, 
                   3, 3, 3, 4, 4, 4, 4, 6, 6, 6 ,6, 3, 3, 6, 6, 6, 6, 6, 
                   6, 3, 3, 3, 3, 4, 4, 4, 6, 3, 3, 3, 3, 6, 6, 3, 3, 3, 3))))
vd <- visualizeDesign(sampleData = sampleData,
                      designFormula = ~ 0 + batch + group, textSize = 4)
cowplot::plot_grid(plotlist = vd$plotlist)
app <- exploreModelMatrix(sampleData = sampleData,
                          designFormula = ~ 0 + batch + group)
if (interactive()) shiny::runApp(app)

vd <- visualizeDesign(sampleData = sampleData,
                      designFormula = ~ 0 + group + batch, textSize = 4)
cowplot::plot_grid(plotlist = vd$plotlist)
```

## Example 5
From https://support.bioconductor.org/p/120850/

```{r, fig.width = 7, fig.height = 16}
(sampleData <- data.frame(
  Response = rep(c("Resistant", "Sensitive"), c(12, 18)),
  Patient = factor(rep(c(1:6, 8, 11:18), each = 2)),
  Treatment = factor(rep(c("pre","post"), 15)), 
  ind.n = factor(rep(c(1:6, 2, 5:12), each = 2))))
vd <- visualizeDesign(sampleData = sampleData,
                      designFormula = ~ Response + Response:ind.n + Response:Treatment,
                      textSize = 3)
cowplot::plot_grid(plotlist = vd$plotlist, ncol = 1)
app <- exploreModelMatrix(sampleData = sampleData,
                          designFormula = ~ Response + Response:ind.n + Response:Treatment)
if (interactive()) shiny::runApp(app)
```

```{r, fig.width = 7}
vd <- visualizeDesign(sampleData = sampleData,
                      designFormula = ~ Treatment + Response, textSize = 4)
cowplot::plot_grid(plotlist = vd$plotlist, ncol = 1)
```

## Example 6
From https://support.bioconductor.org/p/120591/

```{r, fig.width = 7}
(sampleData <- data.frame(
  samples = c("Cell1_HA1", "Cell1_HA2", "Cell1_foxcut11", 
              "Cell1_foxcut12", "Cell2_HA1", "Cell2_HA2", 
              "Cell2_foxcut11", "Cell2_foxcut12"), 
  type = c("Control", "Control", "FOXCUT", "FOXCUT", 
           "Control", "Control", "FOXCUT", "FOXCUT"), 
  replicate = factor(c(1, 2, 1, 2, 3, 4, 3, 4)), 
  celline = factor(c(1, 1, 1, 1, 2, 2, 2, 2))))
sampleData$group <- paste0(sampleData$type, ".", sampleData$celline)
vd <- visualizeDesign(sampleData = sampleData,
                      designFormula = ~ 0 + group + replicate, textSize = 3)
cowplot::plot_grid(plotlist = vd$plotlist, ncol = 1)
```

## Example 7
From https://support.bioconductor.org/p/122641/

```{r, fig.height = 9, fig.width = 7}
(sampleData <- data.frame(individual = factor(rep(1:3, each = 4)), 
                          time = rep(rep(c("before", "after"), each = 2), 3), 
                          exerc = rep(c("trained", "untrained"), 6)))
vd <- visualizeDesign(sampleData = sampleData, 
                      designFormula = ~ individual + time*exerc, textSize = 4)
cowplot::plot_grid(plotlist = vd$plotlist, ncol = 1)
```

```{r, fig.height = 4, fig.width = 7}
sampleData$group <- paste(sampleData$exerc, sampleData$time, sep = ".")
vd <- visualizeDesign(sampleData = sampleData,
                      designFormula = ~ 0 + individual + group, textSize = 4)
cowplot::plot_grid(plotlist = vd$plotlist, ncol = 1)
```

## Example 8
From https://support.bioconductor.org/p/80408/

```{r, fig.height = 5, fig.width = 7}
(sampleData = data.frame(
  condition = factor(rep(c("ctrl_minus", "ctrl_plus", "ko_minus", "ko_plus"), 3)),
  batch = factor(rep(1:6, each = 2))))
vd <- visualizeDesign(sampleData = sampleData,
                      designFormula = ~ 0 + batch + condition, 
                      textSize = 4, lineWidth = 20, 
                      dropCols = "conditionko_minus")
cowplot::plot_grid(plotlist = vd$plotlist, ncol = 1)
```

## Example 9
From https://support.bioconductor.org/p/64480/

```{r, fig.width = 7, fig.height = 7}
(sampleData <- data.frame(
  status = factor(rep(c(0, 1), c(8, 6))),
  rep = factor(c(1, 1, 2, 2, 3, 3, 4, 4, 1, 1, 2, 2, 3, 3)),
  time = factor(rep(c(0, 1), 7))
))
vd <- visualizeDesign(sampleData = sampleData,
                      designFormula = ~ status + time + status:rep + status:time,
                      textSize = 4, 
                      dropCols = "status1:rep4")
cowplot::plot_grid(plotlist = vd$plotlist, ncol = 1)
app <- exploreModelMatrix(sampleData = sampleData, 
                          designFormula = ~ status + time + status:rep + status:time)
if (interactive()) shiny::runApp(app)
```

## Example 10
From https://support.bioconductor.org/p/122764/

```{r, fig.height = 7, fig.width = 7}
(sampleData <- data.frame(
  disease = rep(c("healthy", "disease"), c(6, 9)),
  treatment = rep(c("None", "Drug_A", "Drug_B"), 5),
  patient = factor(rep(c(1:2, 1:3), each = 3))
))
vd <- visualizeDesign(sampleData = sampleData,
                      designFormula = ~ disease + disease:patient + disease:treatment,
                      textSize = 3, 
                      dropCols = "diseasehealthy:patient3")
cowplot::plot_grid(plotlist = vd$plotlist, ncol = 1)
app <- exploreModelMatrix(sampleData = sampleData, 
                          designFormula = ~ disease + disease:patient + disease:treatment)
if (interactive()) shiny::runApp(app)
```

## Session info

```{r}
sessionInfo()
```

